CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-06-11 script/en/chapters/ellipticcurves.tex
be updates, some ecc additions

Mathematicians have been researching elliptic curves for over 100 years. In the
course of time, many lengthy and mathematically complex results have been found
and published in connection with elliptic curves. A mathematician would say that
elliptic curves (or the mathematics behind them) are widely understood. This
research was originally purely mathematical. That is to say, elliptic curves
were investigated, for example, in the mathematical areas of number theory and
algebraic geometry, which are generally highly abstract. Even in the recent
past, elliptic curves played an important role in pure mathematics. In 1993 and
1994, Andrew Wiles\index{Wiles} published mathematical works that triggered enthusiasm far
beyond the specialist audience. In these works, he proved a conjecture put
forward in the 1960's. To put it short, this conjecture was concerned with the
connection between elliptic curves and what are called module forms. That which
is interesting for most people is that the works of Wiles also proved the famous
second theorem of Fermat. Mathematicians had spent centuries (Fermat lived from
1601 to 1665) trying to find a strict proof of this theorem. Understandably,
therefore, Wiles' proof got a good response. Fermat formulated his theorem as
follows (written in the border of a book):

@@

>>>>

Mathematicians have been researching elliptic curves for over 100 years. In the
course of time, many lengthy and mathematically complex results have been found
and published in connection with elliptic curves. A mathematician would say that
elliptic curves (or the mathematics behind them) are widely understood. This
research was originally purely mathematical. That is to say, elliptic curves
were investigated, for example, in the mathematical areas of number theory and
algebraic geometry, which are generally highly abstract. Even in the recent
past, elliptic curves played an important role in pure mathematics. In 1993 and
1994, Andrew Wiles\index{Wiles Andrew} published mathematical works that triggered enthusiasm far
beyond the specialist audience. In these works, he proved a conjecture put
forward in the 1960's. To put it short, this conjecture was concerned with the
connection between elliptic curves and what are called module forms. That which
is interesting for most people is that the works of Wiles also proved the famous
second theorem of Fermat. Mathematicians had spent centuries (Fermat lived from
1601 to 1665) trying to find a strict proof of this theorem. Understandably,
therefore, Wiles' proof got a good response. Fermat formulated his theorem as
follows (written in the border of a book):

@@

One can verify that $+$ defines a group operation on the set $E\cap\{\cal O\}$.
In particular this means that the sum of two points is again a point on the
elliptic curve. How his operation works is geometrically visualized in the
following section.

### 2003-06-11 script/en/chapters/introduction.tex
be update
be updates, some ecc additions

In order to also enable developers/authors to work together independently 
the topics have been split up and for each topic an extra chapter has been 
written which can be read on its own. The later editorial work in TeX added 
cross linkages between different sections and footnotes describing where you
can find the according functions within the CrypTool\index{CrypTool} 
program (see menu tree in appendix A). 
% \hyperlink{appendix-menutree}{(see menu tree in appendix A).}
% in \ref{s:appendix-menutree}   AUFPASSEN, DASS nichts doppelt !!
% \hypertarget{appendix-menutree}{}\label{s:appendix-menutree}
Naturally there are much more interesting topics in mathematics and
cryptography which could be discussed in greater depth -- therefore this
is only one of many ways to do it.

@@

This edition of the script mainly updates the summaries of the following 
topical research areas:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers (generalized Mersenne 
        and Fermat primes) \\ (chap. \ref{spezialzahlentypen}), 
  \item progress in number theory (``Primes in P'') (chap. \ref{PrimesinP}), 
  \item the factorisation of big numbers (TWIRL, RSA-160) 
        (chap. \ref{TWIRLDevice}, \ref{RSA-160}),
  \item progress in cryptanalysis of the AES standard 
        (chap. \ref{NeueAES-Analyse}) and
  \item progress in brute-force attacks against symmetric algorithms 
        (chap. \ref{Brute-force-gegen-Symmetr}).
\end{itemize}

@@

At this point I'd like to thank explicitly 3 people who especially 
contributed to CrypTool\index{CrypTool}. Without their talents and engagement 
CrypTool would not be what it is today:
\vspace{-7pt}
\begin{itemize}
   \item Mr.\ Henrik Koy
   \item Mr.\ J\"org-Cornelius Schneider and
   \item Dr.\ Peer Wichmann.
\end{itemize}
Also I want to thank all the many people not mentioned here for their 
hard work (mostly carried out in their spare time).
\\

>>>>

In order to also enable developers/authors to work together independently 
the topics have been split up and for each topic an extra chapter has been 
written which can be read on its own. The later editorial work in TeX added 
cross linkages between different sections and footnotes describing where you
can find the according functions within the CrypTool\index{CrypTool} 
program \hyperlink{appendix-menutree}{(see menu tree in appendix A).}
% in \ref{s:appendix-menutree}   AUFPASSEN, DASS nichts doppelt !!
% \hypertarget{appendix-menutree}{}\label{s:appendix-menutree}
Naturally there are much more interesting topics in mathematics and
cryptography which could be discussed in greater depth -- therefore this
is only one of many ways to do it.

@@

This edition of the script mainly updates the summaries of the following 
topical research areas:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers (generalized Mersenne 
        and Fermat primes, ``M-40'') 
	\\ (chap. \ref{spezialzahlentypen}, \ref{zahlentyp_mersenne}), 
  \item progress in number theory (``Primes in P'') (chap. \ref{PrimesinP}), 
  \item the factorisation of big numbers (TWIRL, RSA-160)\index{TWIRL device} 
        (chap. \ref{TWIRLDevice}, \ref{RSA-160}),
  \item progress in cryptanalysis of the AES standard 
        (chap. \ref{NeueAES-Analyse}) and
  \item progress in brute-force attacks against symmetric algorithms 
        (chap. \ref{Brute-force-gegen-Symmetr}).
\end{itemize}

@@

At this point I'd like to thank explicitly 3 people who especially 
contributed to CrypTool\index{CrypTool}. Without their talents and engagement 
CrypTool would not be what it is today:
\vspace{-7pt}
\begin{itemize}
   \item Mr.\ Henrik Koy
   \item Mr.\ J\"org Cornelius Schneider and
   \item Dr.\ Peer Wichmann.
\end{itemize}
Also I want to thank all the many people not mentioned here for their 
hard work (mostly carried out in their spare time).
\\

### 2003-06-11 script/en/chapters/numbertheory.tex
be update
be updates, some ecc additions

@@

\bibitem[Bourseau2002]{Bourseau2002} \index{Bourseau 2002} \index{Fox 2002}
    F. Bourseau, D. Fox, C. Thiel, \\
    {\em Vorz\"uge und Grenzen des RSA-Verfahrens},\\
    In: Datenschutz und Datensicherheit (DuD) 26/2002, pp~84-89.

@@

\bibitem[Shamir2003]{Shamir2003} \index{Shamir 2003} 
    Adi Shamir, Eran Tromer, \\
    {\em Factoring Large Numbers with the TWIRL Device}, 
    Januar 2003, \\
    \href{http://www.wisdom.weizmann.ac.il/~tromer/}
    {\texttt{http://www.wisdom.weizmann.ac.il/\~{}tromer/}}.

@@

@@

I would like to take this opportunity to thank 
\begin{itemize}
  \item {Mr.\ Henrik Koy for making many very useful suggestions, 
         for the very constructive proof-reading this article and
         for helping with TeX. \\
	 Mr.\ Koy designed and developed the functions and the complex
	 dialog box of the RSA cryptosystem, which enable you to execute
	 the RSA samples of this article.}
  \item {J\"org-Cornelius Schneider for his TeX support und for the many 
         cases where he helped when facing programming or design problems.}
  \item {Dr. Illies for pointing me to Pari-GP \index{Pari-GP}.}
\end{itemize}

>>>>

The 3-page article in the DuD issue of June 2003 \cite{Weis2003} contains
a very good explanation, how the attack using the Generalized Number Field
Sieve (GNFS) \index{General Number Field Sieve (GNFS)} works and which 
progress is made, to factorize numbers.
At GNFS we can distinguish 2 general steps: 
The sieve step (relation collecting) and the matrix reduction.
Besides the sieve step is highly parallelizable, it dominates the overall
calculation burdon. Shamir and Tromer haven't built a TWIRL device yet,
but the estimated costs of 10 till 50 million Euro (in order to factorise
a 1024-bit number) is not prohibitive for secret agencies or big criminal
organizations, because the ``costs for a single espionage satellite is
estimated e.g. to be several billion USD''. The authors therefore
recommend, to get as soon as possible rid of today used sensible RSA, 
Diffie-Hellman or ElGamal keys up to 1024 bit and to use then keys of at
least 2048 bit length.
The planned TCPA/Palladium hardware \index{Palladium} will use 2048-bit
RSA keys!

@@

\bibitem[Bourseau2002]{Bourseau2002} \index{Bourseau 2002} \index{Fox 2002}
    F. Bourseau, D. Fox, C. Thiel, \\
    {\em Vorz\"uge und Grenzen des RSA-Verfahrens},\\
    In: Datenschutz und Datensicherheit (DuD) 26/2002, pp~84-89 (see www.dud.de).

@@

\bibitem[Shamir2003]{Shamir2003} \index{Shamir 2003} \index{TWIRL device} 
    Adi Shamir, Eran Tromer, \\
    {\em Factoring Large Numbers with the TWIRL Device}, 
    Januar 2003, \\
    \href{http://www.wisdom.weizmann.ac.il/~tromer/}
    {\texttt{http://www.wisdom.weizmann.ac.il/\~{}tromer/}}.

@@

\bibitem[Weis2003]{Weis2003} \index{Weis 2003} \index{Lucks 2003} \index{Bogk 2003}
    R\"udiger Weis, Stefan Lucks, Andreas Bogk, \\
    {\em Sicherheit von 1024 bit RSA-Schl\"usseln gef\"ahrdet},\\
    In: Datenschutz und Datensicherheit (DuD) 6/2003, pp~360-362 (see www.dud.de)\\
    The article explains details about the TWIRL device\index{TWIRL device}.

@@

I would like to take this opportunity to thank 
\begin{itemize}
  \item {Mr.\ Henrik Koy for making many very useful suggestions, 
         for the very constructive proof-reading this article and
         for helping with TeX. \\
	 Mr.\ Koy designed and developed the functions and the complex
	 dialog box of the RSA cryptosystem, which enable you to execute
	 the RSA samples of this article.}
  \item {J\"org Cornelius Schneider for his TeX support und for the many 
         cases where he helped when facing programming or design problems.}
  \item {Dr. Illies for pointing me to Pari-GP \index{Pari-GP}.}
\end{itemize}

### 2003-06-11 script/en/chapters/primes.tex
fixes for beta6
be updates, some ecc additions

\newpage
\section{Prime Numbers}
\hypertarget{Kapitel_2}{}
(Bernhard Esslinger, May 1999, Updates Nov. 2000, Dec. 2001, Feb. 2003)

@@

% --------------------------------------------------------------------------
\hypertarget{MersenneNumbers01}{}
\subsubsection{Special number types -- Mersenne numbers} 
\index{Mersenne!number}

@@

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377; ~
6,972,593; \\13,466,917.
\end{array}
$$
Thus $39$ Mersenne prime numbers are currently known
\index{Prime number!Mersenne}\index{Mersenne!prime number}. 
For the first 38 Mersenne prime numbers we know that this list is complete.
The exponents until the 39th Mersenne prime number have not yet been checked
completely (see chapter 2.5 prime number tests\index{Prime number!test}).

@@

Right now (Feb. 2003) all exponents smaller than $ 6,972,593 $ have been
checked: so we can be certain, that this is really the 38th Mersenne prime
number.
This means there is no other Mersenne prime number between the 37th 
Mersenne prime number and this one.

\vskip +15 pt
\paragraph{M13466917 -- December 2001} \index{Mersenne!prime number!M-39}\mbox{}
\vskip +10 pt

@@

@@

Discovering the 39th Mersenne prime the GIMPS project (Great Internet Mersenne Prime Search)\index{GIMPS} founded in 1996 already discovered for the 5th time the greatest Mersenne number to be proven being prime
({\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}).

@@

Currently the 4 largest known primes are Mersenne prime numbers. Only the 5th biggest known prime is another number type: it belongs to the so called  \hyperlink{generalizedFermatprimes}{generalized Fermat primes}.

@@

The discoverer of M-38 received 50,000 USD from the EFF for discovering 
the first prime with more than 1 million decimal digits. 
The next prize of 100,000 USD offered by EFF is for a proven prime with more
than 10 million decimal digits.
%{\href{http://www.octocad.demon.co.uk/mersenne/prime.htm }{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm}}.

@@

The following enumeration contains the most common ideas for 
``prime number formulae'':
\begin{enumerate}
\item {\em Mersenne numbers}  $f(n) = 2^n - 1$  \quad for $ n $ prime: \\
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 2,000,000 digits are 
    known (M-38\index{M-38}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}
\item $F(k,n) = k \cdot 2^n \pm 1$  for $ n $ prime and $ k $ small primes:\\
   For this generalisation of the Mersenne 
   numbers\index{Mersenne!number!generalized}
   there are (for small $k$)
   also extremely quick prime number tests (see \cite{Knuth1981}). This can
   be performed in practice using software such as the Proths software from
   Yves Gallot\index{Gallot Yves}
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.prothsearch.net/index.html}
    {\tt http://www.prothsearch.net/index.html}}.
\end{itemize}

@@

\textbf{Proven statements / conjectures about primes}

>>>>

\newpage
\section{Prime Numbers}
\hypertarget{Kapitel_2}{}
%(Bernhard Esslinger, May 1999, Updates Nov. 2000, Dec. 2001, Feb. 2003)
(Bernhard Esslinger, May 1999, Updates Nov. 2000, Dec. 2001, June 2003)

@@

% --------------------------------------------------------------------------
\hypertarget{MersenneNumbers01}{}
\subsubsection{Special number types -- Mersenne numbers} 
\index{Mersenne!number}
\label{zahlentyp_mersenne}

@@

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221;
 ~ 3,021,377; ~ 6,972,593; \\
13,466,917.
%13,466,917; ~ xxx,xxx,xxx.
\end{array}
$$
Thus
$39$
%$40$
Mersenne prime numbers are currently known%
\index{Prime number!Mersenne}\index{Mersenne!prime number}. 
For the first 38 Mersenne prime numbers we know that this list is complete.
The exponents until the 39th Mersenne prime number have not yet been checked
completely (see chapter 2.5 prime number tests\index{Prime number!test}).

@@

Right now (Feb. 2003) all exponents smaller than $ 6,972,593 $ have been
tested and double-checked (see home page of the GIMPS project:
{\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}): so we can
be certain, that this is really the 38th Mersenne prime number and that
there are no smaller undiscovered Mersenne primes.

\vskip +15 pt
\paragraph{M13466917 -- December 2001 -- M-39 ?}%
\index{Mersenne!prime number!M-39}\mbox{}
\vskip +10 pt

@@

%\vskip +15 pt
%\paragraph{Mxxxxxxxxxx -- June 2003 2001 -- M-40 ?}%
%\index{Mersenne!prime number!M-40}\mbox{}
%\vskip +10 pt

%This number was discovered as 40th Mersenne prime (and already called M-40,
%despite it has not been proven yet, whether no further Mersenne prime
%numbers between M-39 und Mxxxxxxxxx do indeed exist), $$2^{xx,xxx,xxx}-1,$$
%at June xx, 2003 -- more exactly, the verification of this number,
%found at June 02, 2003 by xxxxxxxxxx, was
%successfully completed. 
%The initiator and project leader George Woltman only announces a found
%Mersenne number, after another double-check confirms that it is prime.
%This number has about xx million decimal digits
%(exactly xx,xxx,xxx digits).

@@

Discovering the 
39th 
%40th 
Mersenne prime the GIMPS project (Great Internet Mersenne Prime Search)\index{GIMPS} founded in 1996 already discovered for the 
5th 
%6th
time the greatest Mersenne number to be proven being prime
({\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}).

@@

Currently
%(June 2003)
the 
4
%5
largest known primes are Mersenne prime numbers. Only the 
5th 
%6th
biggest known prime is another number type: it belongs to the so called  \hyperlink{generalizedFermatprimes}{generalized Fermat primes}.

@@

The discoverer of M-38 received 50,000 USD from the EFF for discovering 
the first prime with more than 1 million decimal digits.

The next prize of 100,000 USD offered by EFF is for a proven prime with more
than 10 million decimal digits.
%{\href{http://www.octocad.demon.co.uk/mersenne/prime.htm }{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm}}.

According to the EFF rules for their prizes they offer in the next stage
150,000 USD for a proven prime with more than 100 million decimal digits.

@@

The following enumeration contains the most common ideas for 
``prime number formulae'':
\begin{enumerate}
\item {\em Mersenne numbers}  $f(n) = 2^n - 1$  \quad for $ n $ prime: \\
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 2,000,000 digits are 
    known (M-38\index{Mersenne!prime number!M-38}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}
\item $F(k,n) = k \cdot 2^n \pm 1$  for $ n $ prime and $ k $ small primes:\\
   For this generalisation of the Mersenne 
   numbers\index{Mersenne!number!generalized}
   there are (for small $k$)
   also extremely quick prime number tests (see \cite{Knuth1981}). This can
   be performed in practice using software such as the Proths software from
   Yves Gallot\index{Gallot Yves}
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.prothsearch.net/index.html}
    {\tt http://www.prothsearch.net/index.html}}.
\end{itemize}

@@

\textbf{Unproven statements / conjectures about primes}
